---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
library(data.table) #install.packages("data.table") if there is no library data.table
library(lattice)
library(ggplot2)
#The reason I use multiple plot library is I am new to R plotting, so for each plot I just googled how to do it(mostly find the answers in http://stackoverflow.com/). Each of the answer uses different plot library.
data=read.csv(unz('activity.zip','activity.csv'))
data <- data.table(data)
summary(data)
```


## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day
```{r}
step_sum_by_date <- as.data.frame(data[, j=list(steps = sum(steps, na.rm = TRUE)),by = date])#total number of steps taken per day
head(step_sum_by_date) # the whole data is very long so we just take a look at first 6 entry
```
2. Make a histogram of the total number of steps taken each day
```{r}
qplot(step_sum_by_date$steps, binwidth = 1000, xlab = "total number of steps taken each day")
#there are some days without any step record, here we simplily leave these days here.
```
3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean(step_sum_by_date$steps)#NOTE: days with zero steps are included
median(step_sum_by_date$steps)
```
## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
step_sum_by_int <- as.data.frame(data[, j=list(steps = mean(steps, na.rm = TRUE)),by = interval])
summary(step_sum_by_int)
plot(step_sum_by_int$interval, step_sum_by_int$steps, type = "l", xlab = "5-min interval", 
    ylab = "Average across all Days", main = "Average number of steps taken")
```
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
step_sum_by_int[which.max(step_sum_by_int$steps),]
```
So the largest number of steps is 206.1698 at interval 835.
## Imputing missing values
1. Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(data))
```
2. Devise a strategy for filling in all of the missing values in the dataset.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
we fill the missing val with average of corresponding interval.
```{r}
new_data <- data
for (i in 1:nrow(data)){
  if (is.na(data$steps[i])){
    interval_val <- data$interval[i]
    steps_val <- step_sum_by_int$steps[which(step_sum_by_int$interval == interval_val)]
    new_data$steps[i] <- steps_val
  }
}
```
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r}
new_step_sum_by_date <- as.data.frame(new_data[, j=list(steps = sum(steps, na.rm = TRUE)),by = date])#total number of steps taken per day
qplot(new_step_sum_by_date$steps, binwidth = 1000, xlab = "total number of steps taken each day with imputated missing value")
mean(new_step_sum_by_date$steps)#NOTE: now there are no zero records
median(new_step_sum_by_date$steps)
```
The mean and median increases because the values with zero now become larger than zero after imputation.However we replace the NA value by the mean of their intervals so the overall shape of data, except in zero point, does not change to much.

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
new_data$date <- as.Date(new_data$date, "%Y-%m-%d")
day <- weekdays(new_data$date)
daylevel <- vector()
for (i in 1:nrow(new_data)) {
    if (day[i] == "Saturday" | day[i] == "Sunday") {
        daylevel[i] <- "Weekend"
    } else {
        daylevel[i] <- "Weekday"
    }
}
new_data$daylevel <- daylevel
new_data$daylevel <- factor(new_data$daylevel)
table(new_data$daylevel)
```
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
data_byweekday <- aggregate(steps ~ interval + daylevel, data=new_data, FUN="mean")
xyplot(steps ~ interval | daylevel, data=data_byweekday, type="l", grid=T, layout=c(1,2), ylab="Average number of steps taken", xlab="5-min. intervals", main="Average  5-min. activity intervals: Weekdays vs. Weekends")
```
